Hippocampal segmentation with MAGeT

### 2013-02-05 paper.Rnw
Updated methods to clarify MAGeT, and add canned description of ADNI1

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is demonstrate the viability of
achieving comparable segmentation accuracy while using significantly fewer
manually created atlases.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion\citep{Collins2010}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follows.  An atlas library and unlabelled MR images are given as input.  Each
atlas image is nonlinearly registered to each unlabelled image, and then each
atlas' labels are propagated via the resulting transformations.  The resulting
labels are fused to produce a single, definitive segmentation by some label
fusion method (for example, by voxel-wise majority voting).

The segmentation approach we propose is best understood as an extension of
MAGeT brain adds a preliminary stage to this process in which the template
library is constructed rather than given as input.  As before, MAGeT brain
accepts an atlas library and unlabelled MR images as input.  Images for the
template library are selected from

To create the template library, labels from each atlas image are propagated to each template
library image via the transformation resulting from a non-linear registration
between pair of images.  As a result, each template library image has a label
from each atlas.  Basic multi-atlas segmentation is then used to produce
segmentations for the entire set of unlabelled images (including those images
used in the template library).

\todo{Note that our algorithm does not specify a registration algorithm or
fusion method\citep{Chakravarty2012}.  Performance on options is tested in our
experiments}

Source code can be found at \url{http://github.com/pipitone/MAGeTbrain}.

@@

<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)

>>>>

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is demonstrate the viability of
achieving comparable segmentation accuracy using the basic multi-atlas schema
and using significantly fewer manually created atlases.

@@

The simplest form of multi-atlas segmentation combines labellings derived from
several atlases by way of label fusion \citep{Heckemann2006}.  We will refer to
this as {\em basic multi-atlas segmentation}.  The primary steps are as
follow are: (1) an atlas library and unlabelled MR images are given as input,
(2) each atlas image is nonlinearly registered to each unlabelled image, (4)
each atlas' labels are propagated via the resulting transformations to the
unlabelled image space, and (5) the resulting labels are fused to produce a
single, definitive segmentation. The particular registration and voting method
used are left unspecified.

The segmentation approach we propose is best understood as an extension of MAGeT
brain adds a preliminary stage to this process in which the template library is
constructed rather than given as input.  As before, MAGeT brain accepts an atlas
library and unlabelled MR subject images as input.  Template library images are
selected from subject images.  The choice of subjects used in the template
library can be made to reflect the neuroanatomy or demographics of the subject
set as a whole (for instance, by sampling equally from cases and controls).

To create the template library, labels from each atlas image are propagated to
each template library image via the transformation resulting from a non-linear
registration between pair of images.  As a result, each template library image
has a label from each atlas.  Basic multi-atlas segmentation is then used to
produce segmentations for the entire set of unlabelled images (including those
images used in the template library). Figure X describes the MAGeT brain
algorithm in pseudocode.

\footnote{Source code can be found at
\url{http://github.com/pipitone/MAGeTbrain}.}

@@

Clinical, demographic and pre-processed T1-weighted MRI used in this paper were
downloaded by the authors from the ADNI1 database (adni.loni.ucla.edu) from
between March 2012 and August 2012.

All images were acquired using 1.5T scanners (General Electric Healthcare,
Philips Medical Systems or Siemens Medical Solutions) at multiple sites from the
ADNI website.  "Representative imaging parameters were TR = 2400ms, TI = 1000ms,
TE = 3.5ms, flip angle = 8°, field of view = 240 × 240mm and 160 sagittal
1.2mm-thick-slices and a 192 × 192 matrix yielding a voxel resolution of 1.25 ×
1.25 × 1.2 mm3, or 180 sagittal 1.2mm-thick-slices with a 256 × 256 matrix
resolution of 1.25 × 1.25 × 1.2 mm3, or 180 sagittal 1.2mm-thick-slices with a
256 × 256 matrix resolution of 1.25 × 1.25 × 1.2 mm3, or 180 sagittal
1.2mm-thick-slices with a 256 × 256 matrix yielding a voxel resolution of 0.94 ×
0.94 × 1.2 mm3." The details of the ADNI MR imaging protocol are described in
\citep{Jack2008a}. The T1-weighted scans that were designated to be the “best”
after quality control were processed using the standard ADNI image processing
pipeline, which included post-acquisition correction of gradient warping
(Jovicich et al., 2006), B1 non-uniformity correction (Narayana et al., 1998)
depending on the scanner and coil type, intensity non-uniformity correction
(Sled et al., 1998) and phantom based scaling correction (Gunter et al., 2006) -
the geometric phantom scan having been acquired with each patient scan.

%We downloaded pre-processed baseline and 12-month repeat volumetric T1-weighted
%MR scans acquired using 1.5T scanners (General Electric Healthcare, Philips
%Medical Systems or Siemens Medical Solutions) at multiple sites from the ADNI
%website.  Representative imaging parameters were TR = 2400ms, TI = 1000ms, TE =
%3.5ms, flip angle = 8°, field of view = 240 × 240mm and 160 sagittal
%1.2mm-thick-slices and a 192 × 192 matrix yielding a voxel resolution of 1.25 ×
%1.25 × 1.2 mm3, or 180 sagittal 1.2mm-thick-slices with a 256 × 256 matrix
%resolution of 1.25 × 1.25 × 1.2 mm3, or 180 sagittal 1.2mm-thick-slices with a
%256 × 256 matrix resolution of 1.25 × 1.25 × 1.2 mm3, or 180 sagittal
%1.2mm-thick-slices with a 256 × 256 matrix yielding a voxel resolution of 0.94 ×
%0.94 × 1.2 mm3. The details of the ADNI MR imaging protocol are described in
%(Jack, Jr. et al., 2008a), and listed on the ADNI website
%(http://www.loni.ucla.edu/ADNI/Research/Cores/). The T1-weighted volumetric
%scans that were designated to be the “best” after quality control were processed
%using the standard ADNI image processing pipeline, which included
%post-acquisition correction of gradient warping (Jovicich et al., 2006), B1
%non-uniformity correction (Narayana et al., 1998) depending on the scanner and
%coil type, intensity non-uniformity correction (Sled et al., 1998) and phantom
%based scaling correction (Gunter et al., 2006) - the geometric phantom scan
%having been acquired with each patient scan.
%
%Clinical and demographic data are shown in Tables 1-4. Table 1 shows the
%demographic data of the 15 randomly-selected subjects (5 control, 5 MCI and 5
%AD) used for method optimisation.  Table 2 shows the demographic data of the 30
%randomly-selected subjects (10 control, 10 MCI and 10 AD) used for method
%validation (note that this subset of 30 subjects was separate from the subset of
%15 subjects used for method optimisation). Table 3 shows the demographic data of
%682 subjects (200 control, 335 MCI and 147 AD) in our full dataset. We also
%sub-divided the MCI subjects into three subgroups (8 reverters, 204 stable and
%123 converters) based on their follow-up clinical diagnoses determined up to 36
%months after baseline.  Table 4 shows the demographic data of the MCI subgroups.
%
<<ADNI-1-screening-demographics, echo=F, results="asis",cache=FALSE>>=
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 + MMSE, 
               data = adnimerge, subset = VISCODE == 'bl',   
               method = "reverse", test=FALSE, overall=TRUE)
